Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-04 paper/paper.tex
a few more tweaks to Discussion

\section{Discussion}
We have determined the full sequences of the viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring the abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell transcriptomic techniques.
To make this information complete, we have coupled a standard single-cell transcriptomic technique to long-read PacBio sequencing of viral genes, a strategy analogous to that recently used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

This viral genetic information turns out to be extremely important for understanding infection outcome and innate-immune induction.
Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, a large majority of cells are infected by virions that have mutations or fail to express a gene.
Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experimental approach that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.

It turns out that these idiosyncratic viral defects explain much of the heterogeneity in viral transcriptional burden that has been observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
If we examine only cells infected by truly wildtype virions, we see that the heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).

Crucially, viral defects also play an important role in determining if infected cells trigger an innate-immune response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---likely act by impairing NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
In fact, given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
There may be several reasons: first, we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; second, our experiments preferentially captured cells with higher viral transcriptional load; and third, most prior studies have used techniques that can detect large deletions but not subtle point mutations.
But we emphasize that the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it certainly remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

Our results also indicate that different types of viral defects induce IFN by somewhat different processes.
For instance, higher viral gene transcription is associated with IFN induction only in cells that are infected by virions with deficiencies in NS.
This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to cell-to-cell variability in the expression of anti-viral genes.
Larger-scale studies similar to the one performed here should be useful for further determining the extent that virions with different defects cause cells to trigger distinct transcriptional programs.

\jdbcomment{But results also show that non-viral factors are clearly important.}

>>>>

\section{Discussion}
We have determined the full sequences of all viral genes expressed in single influenza-infected cells, and examined how viral mutations and gene-expression defects associate with infection outcome in each cell.
Methodologically, our major advance is to measure the \emph{genotypes} of viruses in addition to measuring abundance of viral components (i.e., transcripts, proteins, or progeny virions) as has been done by prior single-cell studies~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, zhu2009growth, schulte2014single, akpinar2016high, heldt2015single, brooke2013most}.
Our method builds on the observations by \citet{saikia2018simultaneous} and \citet{zanini2018virus} that fragmentary viral genetic information can be obtained using more standard single-cell transcriptomic techniques.
To make this information complete, we have coupled a single-cell transcriptomic technique with long-read PacBio sequencing of viral genes, a strategy analogous to that used by \cite{gupta2018single} to obtain full-length isoforms of some cellular genes in single cells.

This viral genetic information turns out to be crucial for understanding infection outcome and innate-immune induction.
Despite the fact that we used a low-passage viral stock generated from plasmids encoding a ``wildtype'' influenza genome, most cells are infected by virions that have mutations or fail to express a gene.
Although our study is certainly not the first to note that influenza virus has a high mutation rate~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel} and sometimes fails to express genes~\citep{brooke2013most, heldt2015single, dou2017analysis, russell2018extreme}, it is the first to directly observe the full spectrum of these defects across single cells.
Simple visual inspection of \FIG{genotypes} makes it glaringly obvious how any experiment that does not directly sequence the viral genes in single cells is really averaging across a vast number of hidden viral defects.

It turns out that these viral defects explain much of the heterogeneity in viral transcriptional burden observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}.
If we examine only cells infected by truly wildtype virions, we see that heterogeneity in viral transcriptional burden is substantially reduced (\FIG{mutations}A).

Viral defects also help determine if an infected cell triggers an innate-immune response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising that we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI~\citep[conditions that better reflect the early stages of real human influenza infections;][]{mccrone2018stochastic, xue2018reconciling, varble2014influenza}; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
But the relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

Our results also indicate that different viral defects induce IFN by somewhat distinct processes.
For instance, higher viral transcription is associated with IFN induction only in cells infected by virions with deficiencies in NS.
This finding is consistent with the suggestion by \citet{sjaastad2018distinct} that viral genetic variation contributes to cell-to-cell variability in the expression of anti-viral genes.
Larger-scale studies similar to the one performed here should be useful for further elucidating the extent that virions with different defects cause cells to launch distinct transcriptional programs.

But although viral variation is clearly important, it still only partially explains heterogeneity among influenza-infected cells.
We find substantial breadth in viral transcriptional burden and occasional IFN induction even among cells infected with fully wildtype virions.
Of course, our single-cell virus sequencing only calls mutations present in a cell at relatively high frequency, so there could be low-frequency but dominant mutations that affect the behavior of cells expressing predominantly wildtype viral genes.
However, we suspect that pure stochasticity or pre-existing cellular states also affect the outcome of viral infection, since heterogeneity in IFN induction is present even among cells treated with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single, bhushal2017cell}.
